---
title: "TEfastR: fast gene/TE counting, from R"
author: "Sulev Koks"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: kable
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What this is

[TEtranscripts](https://github.com/mhammell-laboratory/TEtranscripts) quantifies
genes and transposable elements (TEs) from RNA-seq, redistributing
multi-mapped reads across TE instances with an EM algorithm. Its counting
stage reads a BAM one alignment at a time through `pysam` and, for every
read, runs a Python-level loop to build CIGAR-derived intervals and query an
interval index.

`TEfastR` re-implements that same counting stage in R, but vectorized:

* **One bulk BAM read**, via `GenomicAlignments::readGAlignments()`, instead
  of a per-alignment loop.
* **Bulk overlap queries**, via `GenomicRanges::findOverlaps()`, run once
  for *all* reads against the annotation at once, instead of one query per
  read.
* **A sparse-matrix EM step** (`Matrix` package) for redistributing
  multi-mapped reads across TE instances, instead of a pure-Python loop over
  every multi-mapped alignment on every iteration.

It reproduces TEtranscripts' assignment rules: unique vs. multi-mapped
classification, `--stranded` matching, proportional ambiguity resolution for
reads overlapping more than one gene or TE instance, and the same
SQUAREM-accelerated EM used by TEtranscripts' own `EMAlgorithm.py`.

### Validation against real data

`tests/testthat/test-count-te.R` cross-checks TEfastR against TEtranscripts'
own Python counting core (`TEToolkit.Counting.count_transcript_abundance`)
on synthetic single-end and paired-end fixtures: unique gene/TE assignment,
locus-level ambiguity, a genuinely multi-mapped read (exercising the EM
step), paired-end pairing including singletons, all three `--stranded`
modes, and the fragment-length estimate used for TE effective-length
correction.

That synthetic suite is necessary but not sufficient -- several of the bugs
below only showed up on real data, where annotation patterns (many
overlapping transcript isoforms, thousands of near-identical repeat copies)
are far messier than anything a small hand-built fixture exercises. On a
real 44.5-million-alignment paired-end BAM (human, GENCODE v50 +
RepeatMasker), TEfastR completed GTF loading, counting, and 100 EM
iterations in **7.3 minutes** on a single core (~54GB peak RSS); per-feature
counts on a chromosome-14 subset (3,894 genes and TE elements, 6.4M
alignments) now correlate with the real Python `TEcount` at **r = 0.99999**,
with 91% of features agreeing to within 1%. Getting there surfaced four real
bugs, since fixed (see `NEWS`/git history for the fixes; each has a
regression test):

* A unique read whose single locus overlaps multiple genes (e.g. two
  transcript isoforms' exons) must be split evenly and *immediately*.
  TEtranscripts does **not** defer this to its proportional
  ambiguity-resolution step the way it does for the equivalent single-locus,
  multiple-TE-instance case. Deferring it (this package's original
  approach) lets a gene with any confidently-assigned reads elsewhere
  silently absorb every read that should have been split with a neighbor
  that has none -- found via a gene nested inside a multi-isoform host
  gene's shared terminal exon, where one gene absorbed 2x the read pair's
  true combined total while the other got zero.
* Similarly, a genuinely multi-mapped read that hits no TE anywhere falls
  back to gene assignment as *one* deferred entry per **read** (pooling
  every one of that read's loci, weighted `1/(number of loci)`), not one
  entry per ambiguous locus.
* TEtranscripts' naive prior for a multi-mapped read's TE candidates divides
  by the number of loci that had *at least one TE hit*, not the read's total
  locus count -- a read can multi-map to a mix of TE-annotated and
  TE-desert loci, and only the former count toward the divisor.
* TEtranscripts' EM step is handed an all-zero placeholder for `uniq_counts`
  internally; the real unique counts are added to the EM's output only
  afterward. Feeding the real counts into the EM's own effective-length
  calculation instead skews its starting point so TEs with many confident
  unique reads absorb ambiguous multi-mapped reads that belong elsewhere --
  this was the dominant remaining source of error on repeat-dense regions
  until fixed.
* The fragment-length estimate that drives TE effective-length correction
  must come from unique reads only, exclude proper pairs with an insert
  size over `maxL` (default 500, matching TEtranscripts' `--maxL`), and be
  floor-divided (integer division, as in the Python `//` TEtranscripts
  uses) rather than averaged as a float. For a short TE instance, the
  fractional-vs-floored difference -- or a single unfiltered outlier
  insert size -- can be enough to flip its effective length across zero,
  excluding it from EM entirely and handing its reads to a different,
  longer element instead.

Residual differences after all four fixes are small (the 9% of features
outside 1% agreement) and consistent with the CIGAR-handling difference
noted below, not a further correctness bug -- but treat any TEfastR output
you're about to publish on as worth a spot-check against TEtranscripts on a
subset first, the same way you'd sanity-check any reimplementation of a
published method.

### Known differences from TEtranscripts

* CIGAR `"D"` (deletion) is kept within one contiguous aligned block here,
  via `GenomicAlignments::grglist()`. TEtranscripts's `fetch_exon()` starts a
  new block at every CIGAR operation, splitting a read into extra tiny
  fragments around small deletions. Only `"N"` (splice/intron) breaks a
  block in both implementations -- this only matters for reads overlapping a
  short indel very close to a feature boundary.
* A BAM is assumed to be consistently single-end or paired-end, checked once
  from the first alignment, rather than decided per read.
* No `--sortByPos` equivalent is needed: reads are grouped by name via a
  hash join, not file adjacency, so a coordinate-sorted, indexed BAM is read
  directly -- this also means TEfastR skips the name-sort pass TEtranscripts
  needs for coordinate-sorted input.

## Installation

```{r, eval = FALSE}
# from the TEtranscripts repository checkout
install.packages("TEfastR", repos = NULL, type = "source")
# or, for iterative development:
devtools::install("path/to/TEfastR")
```

## Quick start

Build the gene and TE annotation indexes once -- they don't depend on any
particular BAM, so build them once per project and reuse them across every
sample.

```{r setup}
library(TEfastR)

te_gtf <- system.file("extdata", "te.gtf", package = "TEfastR")
gene_gtf <- system.file("extdata", "gene.gtf", package = "TEfastR")
bam <- system.file("extdata", "test.bam", package = "TEfastR")

te_index <- build_te_annotation(te_gtf)
gene_index <- build_gene_annotation(gene_gtf)

te_index$n          # number of TE instances
length(gene_index$genes)
```

Then count one BAM file:

```{r}
result <- count_te(bam, gene_index, te_index, stranded = "no", mode = "multi", iteration = 100)
result
result$gene_counts
result$te_element_counts
```

`count_te()` returns a `TEfastR_result` list with `gene_counts`,
`te_instance_counts` (per TE instance, before grouping into elements),
`te_element_counts` (TE instances summed by `gene_id:family_id:class_id`,
matching TEtranscripts' own output granularity), and `stats` (read
classification counts and the estimated fragment length used for the EM
step's effective-length correction).

Convert to a plain data frame, or write TEtranscripts' own `.cntTable`
format directly:

```{r}
head(as.data.frame(result))
```

```{r, eval = FALSE}
write_count_table(result, "sample1.cntTable")
```

## Counting several samples and handing off to DESeq2

`count_te_batch()` counts a set of BAM files and assembles a feature x
sample matrix -- the same shape `DESeq2::DESeqDataSetFromMatrix()` expects,
so the whole pipeline can stay inside one R session instead of shelling out
to TEtranscripts' generated R script:

```{r, eval = FALSE}
bams <- c(
  treatment1 = "treatment_rep1.bam", treatment2 = "treatment_rep2.bam",
  control1   = "control_rep1.bam",   control2   = "control_rep2.bam"
)

counts <- count_te_batch(bams, gene_index, te_index, stranded = "no", mode = "multi")

library(DESeq2)
sample_info <- data.frame(
  group = factor(c("treatment", "treatment", "control", "control")),
  row.names = colnames(counts)
)
dds <- DESeqDataSetFromMatrix(countData = round(counts), colData = sample_info, design = ~ group)
dds$group <- relevel(dds$group, ref = "control")
dds <- DESeq(dds)
results(dds)
```

(`round()` is needed because TE counts are fractional after EM
redistribution -- the same rounding TEtranscripts' own generated R script
applies before handing counts to DESeq2.)

## Command-line usage

For use outside R/RStudio, a CLI wrapper ships in `inst/scripts`, mirroring
TEtranscripts' own `TEcount`:

```{r, eval = FALSE}
system.file("scripts", "run_te_count.R", package = "TEfastR")
```

```sh
Rscript "$(Rscript -e 'cat(system.file("scripts","run_te_count.R",package="TEfastR"))')" \
  --bam sample.bam --gtf genes.gtf --te te.gtf \
  --stranded no --mode multi --iteration 100 --out sample.cntTable
```

## Reference

* `build_gene_annotation()` / `build_te_annotation()` -- parse GTFs once.
* `count_te()` -- count one BAM against a pair of annotation indexes.
* `count_te_batch()` -- count several BAMs into one count matrix.
* `write_count_table()` / `as.data.frame.TEfastR_result()` -- get counts out.
